---
title: "How to Embed External micro-Flipbooks in Bookdown"
description: |
  After you have created micro-flipbooks in your miniflipbook project, how do you embed them in Rmarkdown and bookdown projects?
  
author:
  - name: Peter Higgins
    url: https://higgi13425.github.io/medical_r/
date: 10-20-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

## How to micro-flipbook
Let's assume you followed along in the micro-flipbooks blog post, 

https://higgi13425.github.io/medical_r/posts/2020-10-14-micro-flipbooks/

and you now have an RStudio project with at least one micro-flipbook, which has been pushed to Github, and is now viewable at your github.io site, much like this one:

## Example micro-flipbook

Try it out - click on the link below, then use the left and right arrows to go forward and back in the xaringan slides.

https://higgi13425.github.io/mini_flipbooks/ttest_flipbook.html#1 

So now that this is on the web as a viewable flipbook, how are you going to embed it in your Rmarkdown or bookdown project?

Some great ideas on how to do this properly (not the slapdash [but quick] way that I do it) from the source of all flipbooks, Evangeline Reynolds, are here:
https://evangelinereynolds.netlify.app/post/embedding-flipbook-mini-in-html-document/

Also, lots more flipbook knowledge/options here:

https://evangelinereynolds.netlify.app/#posts


## Fun with iframes!

Yes, in addition to stealing the brilliant idea of flipbooks from Evangeline Reynolds (@evamaerey), we are going to steal a trick or three from the _iframe_ knowledge of Desiree DeLeon's (@docossyle) blog post on how to use iframes to embed _learnr_ apps.

For full details on how to do this properly, refer to Desiree's blog post at:

http://desiree.rbind.io/post/2020/learnr-iframes/

## Down to the details

This is actually not that hard.

Once you have working HTML at github.io with a web address, you are in business.

Just insert some _iframe_ code between these comment fences, as in the code chunk below.
This is actually *easier* than inserting a _learnr_ app - with no resizing messes.

Note that it is essentially two html tags, one to open the iframe: `<iframe>` with a bunch of options,

and then one to close the iframe: `</iframe>`

The style argument is basically some inline CSS styling.

Note that if you have multiple iframes in your HTML, each one needs a unique id.

You can set the width and height (these seemed to work for me).

Then set the source (src) to the link to github.io, with the first xaringan slide (#1) as the starting point.

Then usually set scrolling to "no". Your call. 

```{r, eval FALSE, echo=TRUE, error=TRUE, warning=FALSE, message=FALSE}
<!---FLIPBOOK EX 1--> 

<iframe style="margin:0 auto; border: solid black;" 
id="myIframe8" 
width = "763" height = "432" src="https://higgi13425.github.io/mini_flipbooks/ttest_flipbook.html#1" 
scrolling="no"></iframe> 

<!-------------> 

```

Then, build your website or bookdown book with this chunk in it.

It may be helpful to preface the first microflipbook with a bit of orientation for new users. Something like:

See the example flipbook below. You can click on it and move forward or backward in the slides with the left and right arrows to see what each line of code actually does.
