---
title: "Building a Data Package"
description: |
  For building a package that is largely about making datasets readily available in R.
author:
  - name: Peter Higgins
    url: https://example.com/norajones
date: 10-18-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Getting Started on a Data Package

Most blog posts about building packages are about bundling functions that are related to each other into a single, thematic package. This is a common usage of packages, but at times you may want to share some common data sets with users in the form of a package. This is not as well-documented, but can be valuable.
<br><br>
A few examples of data packages include:

- babynames
- fueleconomy
- nasaweather
- nycflights13
- usdanutrients
<br><br>
To get started, you will want to follow the guidance in the R Packages Book at: https://r-pkgs.org/whole-game.html.
This blog post will be a quick getting started guide, relative to the big R Packages book (but use RP as a reference if you get stuck).

## Setup
You want to have a new project open in RStudio, with the project named something like {datapackage} (pick a more appropriate name for your particular package). You then want to load (library commands in the Console) some critical packages including {devtools}, {usethis}, and {roxygen2}.

- `library(devtools)`
- `library(usethis)`
- `library(roxygen2)`

## Creating the Package
Then use the *create_package()* function to create your new package {datapackage}, by running this function in the Console (pick a better name than datapackage):

`create_package("datapackage")`

This will set up your directories and folders in the current project to help you create a package.

## Versioning with Git and Github

If you have never used git before for version control, this would be a good time to install and initialize git on your computer, using the guidance in Chapter 6 of **Happy Git with R**, at: 
https://happygitwithr.com/install-git.html 

Details on how to configure git are in chapter 7, at https://happygitwithr.com/hello-git.html

Now make a git repository to track your versions, by running this in the Console: 

`use_git()`

Go ahead and commit files when prompted to do so.

If you have never used GitHub before for version control and collaboration, this would be a good time set up a Github account at www.github.com and to add a PAT (Personal Access Token) to your computer, using the guidance in Chapter 4 of **Happy Git with R**, at: 
https://happygitwithr.com/github-acct.html

Details on PATs (you want one, trust me) are found in Appendix B, at: https://happygitwithr.com/github-pat.html

## Edit the Description
Look in your Files tab, open the DESCRIPTION file, and edit:

- Make yourself the author (and add collaborators!)
- add a title
- add a longer Description - who is it for, why would they want this data, what is a use case?

## Add a License
Mke your life easy, and 
run 
`use_mit_license()`
in the Console

## Add data
In the Files tab in RStudio, create a `raw-data` folder in your project. Add the raw files (csv, excel, etc.) that you want to turn into your accessible datasets. 
<br>
Now create a separate `data` folder in your project. 

In the `raw-data` folder, create a new R script named `prep_data.R`.
In this file, wrangle, filter, or otherwise prepare your datasets. Let's pretend you have a datafile that you have wrangled from Excel, and it is now a dataframe named `sportsdata`.

Now you want to save this datasets in `*.rda` format, as `sportsdata.rda`. Pick the name carefully, as you will have to live with it for a long time.

In this script, do this saving step with with
`saveRDS(sportsdata, "data-raw/sportsdata.Rdata")`

Then **use** the data (put the data in the proper format in the `data` folder) with the following line in your script:

`usethis::use_data(sportsdata, overwrite = TRUE)`

## Installing the Package

## Deal with Dependencies
Do you need to use other packages to run functions (most of the time not, for a data package)?

On the off chance you do, you will want to run:
`use_package("packagename")`

As in, if you need readr to read in some data, run in the Console:

`use_package('readr')`

## Set up for push to GitHub
Run in the Console:

`use_github()`
This will link your project to a Github repository, so that you can share your datapackage with the world.

## Basic documentation
Create an R script in the `R/` folder, named something like `datapackage.R`.

In this file, you can add roxygen2 documentation, which looks like a long list of funny-looking comments (they start with `#'` rather than just a hashtag).

Typically you include: 

- a short description, in a sentence or two.
- after the @format statement, details on the format (dataframe) and the dimensions (how many rows, how many columns), and a short description of each variable (item), with levels or units; and the data type.
- after the @source statement, a description of where the data came from, often including a link.

This looks something like:

```{r, echo=TRUE, eval=FALSE}
#' Baby names.
#'
#' Full baby name data provided by the SSA. This includes all names with at
#' least 5 uses.
#'
#' Lifetables
#'
#' Cohort life tables data as provided by SSA.
#'
#' @format A data frame with nine variables:
#' \describe{
#' \item{\code{x}}{age in years}
#' \item{\code{qx}}{probability of death at age \code{x}}
#' \item{\code{lx}}{number of survivors, of birth cohort of 100,000, at next integral age}
#' \item{\code{dx}}{number of deaths that would occur between integral ages}
#' \item{\code{Lx}}{Number of person-years lived between \code{x} and \code{x+1}}
#' \item{\code{Tx}}{Total number of person-years lived beyond age \code{x}}
#' \item{\code{ex}}{Average number of years of life remaining for members of cohort alive at age \code{x}}
#' \item{\code{sex}}{Sex}
#' \item{\code{year}}{Year}
#' }
#'
#' For further details, see \url{http://www.ssa.gov/oact/NOTES/as120/LifeTables_Body.html#wp1168594}
#'
"lifetables"
```

Note that use can use Rmarkdown formatting for this file (and for similar files throughout your package) if you add the following line to your package DESCRIPTION file: <br>
`Roxygen: list(markdown = TRUE)`

To build your roxygen2 document into a proper help file, you need to run
`devtools::document()`.  This will use roxygen2 to convert this `*.R` script file to a `*.Rd` file in the `man` folder. This provides the "manual" for your data.

To build this into the package, re-build the package, with Cmd-Shift-B (Mac) or Ctrl-Shift-B (Windows), or click on the `Install and Restart` button under the Build tab in RStudio. This will rebuild the package, restart R, install the package, and load (library) the package.

Now you should be able to call from the console:    

`?dataset` or `help(dataset)`

and see your documentation.

Then save the files and make a Git commit and then push the updated package up to your Github repository.


## Advanced documentation
*Description Document* You can write a long description of the provenance of the data, when and how it was collected, where it was published in a journal, etc. (go for it, provide all the details!) in a Rmarkdown file (named something like datapackage_desc.Rmd), then knit this to pdf. When you push your project/package to Github, your pdfs show up properly formatted. You can then take the link to the pdf and put it in your README file, so that anyone can read up on your datasets.
<br><br>
*Codebook Document* You can create a nice table of each variable, a longer description, units of measurement, levels/category codes and what they mean, in a nicely formatted Rmd table. Create a Rmarkdown file (named something like datapackage_codebook.Rmd), then knit this to pdf. When you push your project/package to Github, your pdfs show up properly formatted. You can then take the link to the pdf and put it in your README file, so that anyone can know what each variable in your dataset is about.

## Add a README
Just run the following in the Console:

`use_readme_rmd()`

Then open README.Rmd and edit this.
This is a good place to put a description of the package, who it is intended for, and some likely use cases. 

This is a helpful place to let users, especially new users, know how to load your package, and how to access the datasets. Show an example or two.

It is also a good place to add a table with the names of the datasets and links to the descriptions and codebooks.


## Check and install

`check()`

`install()`

## Build

The Build tab in RStudio will let you `Build Package` which will also restart R.

This will give you a clean slate and re-load (library) your datapackage.

Once you are happy with the datapackage, commit and push to Github. Now anyone can use you datapackage.