---
title: "micro-flipbooks"
description: |
  How to make embeddable micro-flipbooks (single topic example) for embedding in bookdown HTML in e-books
author:
  - name: Peter Higgins
    url: https://higgi13425.github.io/medical_r/
date: 10-14-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(flipbookr)
library(xaringanthemer)
library(tidyverse)
library(infer)
library(medicaldata)
prostate <- medicaldata::blood_storage %>% as_tibble()
```

## How to make an embeddable micro-flipbook

Start with picking a topic or function you want to teach.

Draft, on paper, a rough outline of the topic and the examples you want to show, building in complexity.

Open the mini-flipbooks project in RStudio. You can fork this from https://github.com/higgi13425/mini_flipbooks if you would like.

### New File
Open a new file -
File/New/Rmarkdown/From Template/Ninja Themed Presentation

(Note: you could also just copy/paste in a working example, like **across_microflip_1** -
this will include YAML edits, setup packages, etc.)

### Name it
Then give it a name, like
<br>
function-name_microflip_#
(increment # for each additional example)

### Delete Lots of Stuff
Then delete everything from line 30 - line 576 if you are using the Ninja template.

### Edit the YAML
Update

- title
- edit or remove subtitle
- author
- add one line to nature:
  - `ratio: 16:9`

### Edit the Setup chunk
Edit opts_chunk$set to include
```{r}
knitr::opts_chunk$set(fig.width = 6, message = FALSE, warning = FALSE, comment = "", cache = FALSE, fig.retina = 3)
```
and delete
options(htmltools.dir.version = FALSE)

Then Add

- `library(flipbookr)`
- `library(xaringan)`
- `library(xaringanthemer)`
- `library(medicaldata)` or other data source
- other libraries that are needed for functions
- load any data files needed into a named dataframe, like <br>
<br>
prostate <- medicaldata::blood_storage %>% as_tibble()


### Edit the xaringan-themer chunk
For a consistent style, I use:

```{r, echo=TRUE, eval = FALSE}
xaringanthemer::style_duo_accent(
  primary_color = "#00274C",
  secondary_color = "#FFCB05",
  header_font_google = google_font("Lato"),
  text_font_google   = google_font("Manrope"),
  code_font_google   = google_font("Roboto Mono"),
  base_font_size = "20px",
   code_font_size = "2rem",
)   
```

You can make your own version.

### Start adding headers

3 hashtags - title line on slide <br>
4 hashtags - subhead text <br>
Can also add body text to explain what you are going to be showing.

then add a divider to start a new slide with 3 horizontal dashes on a single line. <br>
This next slide will be your flipbook slide

### Add your named chunk of R code
- give it a name, like *ttest1*
- use include = FALSE
- Example:

```{r ttest1, echo = TRUE, eval = FALSE}
# start with data
prostate %>%
# apply t_test from infer
# with correct arguments
  t_test(TVol ~ AA, 
         order = c("0", "1"))
# Interpreting the output:
# The t statistic is first
# followed by degrees of freedom
# then the p value
# the default alternative: two.sided
# then the confidence bounds
# output is a tibble so that it is 
# easy to use these results
```

<br>
Note that explanatory material is put in comments with a hashtag - can be in-between code lines, or at the end

Then **reveal** your code line by line with  inline r code using the `chunk_reveal` function:

```{r ttest1-reveal, echo = TRUE, eval = FALSE}
chunk_reveal("ttest1", 
    break_type = "auto", 
    title = "Tidy version from the {infer} package")`
```

<br>
Note that this uses the name of the chunk above, <br>
sets the break_type to "auto", and <br>
gives the microflipbook slide itself a title.

Then (if you want more explanatory material) insert a new slide by adding 3 dashes alone on a single line

Add the explanatory text with Rmd formatting.

Then (if desired) add a css chunk (example below) to adjust your code font - can adjust line height or font size.

This css chunk does not have to be on a separate line

```{css, eval = FALSE, echo = TRUE}
.remark-code{line-height: 1.5; font-size: 80%}
```
