---
title: "Functions and Functionals: TidyEval Fun"
description: |
  Two approaches to use tidyeval to work in both individyal functions and with functionals like map and map2.
author:
  - name: Peter Higgins
    url: {}
date: 2022-08-26
output:
  distill::distill_article:
    self_contained: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(palmerpenguins)
library(purrr)
```

```{r}
knitr::include_graphics(
  here("_posts/2023-08-26-functions-and-functionals-tidyeval-fun/tidyeval.png"))
```

Tidyeval is a way to use the tidyverse in functions that you can write
yourself without resorting to base R, and in functionals (functions like *map*, *map2*, and *pmap*
within the {purrr} package).

But tidyeval can be confusing, in part because it is still evolving, and in part because there are often multiple ways to do the same thing.

As a side note, there are many, many, many (possibly too many) ways to use tidyeval, with ensyms and enquos and !! and !!! to evaluate expressions, but these two 3-part approaches get me through the day.

## Approach 1 - .data prefix the vars and Quote varnames in vectors

Let's start with one approach - by writing a function with the .data prefix - the .data prefix is telling R that the in-dataframe variable names that follow in [[double brackets]] come from INSIDE a dataframe. Let's start with a function named `plot`, to make a ggplot.

```{r}
plot <- function(data, x, y){ 
  ggplot(data) + aes(.data[[x]],
.data[[y]]) + geom_jitter() 
  }
```

Now let's test this function.
Note that we have to QUOTE
within-dataframe variable names when you use the .data prefix in the function

```{r}
plot(penguins, 'bill_length_mm', 'bill_depth_mm')
```

OK, that works fine for a single use. But what if you want to use `functionals` - the functions from the {purrr} package that let you do this across a vector of inputs in one function call?

Let's set this up with vectors for multiple x values and multiple y values.
Four values for both x and y.
And then we will make a dataframe of all possible combinations (16 combos = 4*4) called analysis_list.

```{r}
xvars <- c('bill_length_mm', 'bill_depth_mm', 'flipper_length_mm',
'body_mass_g')

yvars <- c('sex', 'year', 'island', 'species')

analysis_list <- crossing(xvars, yvars) # all combinations of x and y

xvars
yvars
analysis_list
```

OK, that looks good for our 3 possible inputs to a functional.

Let's try this out by putting xvars and yvars into the functional, _map2_.

```{r}
map2(.x = xvars, .y = yvars, .f = plot, data = penguins, .progress =
TRUE) #single pairs of x, y
```
OK, that worked, and it paired up each value of xvar with the corresponding value of yvar to make 4 plots.

Now let's try all possible (16) combinations.

```{r}
map2(.x =analysis_list$xvars, .y = analysis_list$yvars, .f = plot, data
= penguins, .progress = TRUE) # all combos of x, y
```

This works - we can use the 

1) .data[[var]] prefix in the function for in-dataframe variable names, and
2) quote the variables when we put them into a vector for the functionals to iterate over.
3) don't do anything extra in functionals like _map_ or _map2_.

## Approach 2 - Embrace vars and Syms vectors

NOTE that there is another distinct approach, which does not combine well with the .data prefix.

This is to 

1) embrace (curly-curly) the {{in-dataframe variable names}} in the function with the {{var}} wrapping, and
2) do NOT use any quotes for in-dataframe variable names when using the stand-alone function, and
3) turn the unquoted variable names in the vectors input into the functionals into symbols with the _syms_ function.

Let's see how this works. We will make a new function, plot2, with curly-curly around x and y (the in-dataframe variables), and just for variety, we will use geom_boxplot.

```{r}
plot2 <- function(data, x, y){ 
  ggplot(data) + aes({{x}}, {{y}}) +
    geom_boxplot() 
  }
```

Let's test out the stand-alone function (note NO manual QUOTING of variables)

```{r}
plot2(penguins, bill_length_mm, bill_depth_mm)
```

To reiterate, you do NOT have to quote variable names when using a function made with embrace/curly-curly

But this gets wonky with functionals like map2 - you get wonky plots - it 'works' (no error thrown), but the plots are wrong.
See below, tested with single pairs of x and y

```{r}
map2(.x = xvars, .y = yvars, .f = plot2, data = penguins, .progress =
TRUE) 
```

However, embrace/curly-curly works OK with functionals if you then (within the functional) convert both the xvars and yvars into symbols with the _syms_ function, using single pairs of x and y.

```{r}
map2(.x = syms(xvars), 
     .y = syms(yvars), 
     .f = plot2, 
     data = penguins) 
```

The syms approach also works for the analysis list of all combinations, to give you 16 plots of all combos.

```{r}
map2(.x = syms(analysis_list$xvars), 
     .y = syms(analysis_list$yvars), 
     .f = plot2, 
     data = penguins) 
```

## Summary

So two approaches to tidyeval can work, but are not mutually compatible

### Approach 1

1) use the .data[[var]] prefix for each in-dataframe variable name when you build your function
2) QUOTE the in-dataframe variables when you use them in the function
3) no extras added when you use functionals like _map2_.

### Approach 

1) Wrap each {{in-dataframe variable name}} with curly-curly when you build your function
2) NO NEED to quote the in-dataframe variables when you use them in the function
3) Then wrap the vectors of in-dataframe variables in the _syms_ function within functionals like _map2_.

## Other Resources

1. Tidyeval videos https://www.rstudio.com/tags/tidyeval/

2. Schloss lab Intro to Tidyeval https://github.com/SchlossLab/tidy-eval

3. Mile McBain Friendly eval interface https://github.com/MilesMcBain/friendlyeval


