---
title: "How to Post"
description: |
  A reference for future me, on how to create posts.
author:
  - name: Peter Higgins
    url: https://example.com/norajones
date: 10-14-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

For the distill blog **Medical R**,
based on the repo at https://github.com/higgi13425/medical_r

in the R project **medical_r**

and viewable on the web at https://higgi13425.github.io/medical_r/


## 1. Create a new post

run in the Console in RStudio <br>
`distill::create_post("title-of-post")`

this creates a new Rmd document for editing, inside the `_posts` folder, and within its own _date-title_ specific folder inside the `_posts` folder. You can now edit this Rmd, including the title in the YAML header. 


## 2. Editing

All the usual Rmarkdown features, including embedding code chunks, images, videos, etc.<br>

https://rmarkdown.rstudio.com/authoring_basics.html 

### Adding Images to your Post

1. load the `here()` package in the setup chunk.
2. save the image you want with a short name in the same post folder as your blog post, adjacent to the Rmd file. Note the image name and extension.
3. Insert a code chunk that looks like this, but with echo = FALSE, and eval = TRUE
```{r, echo=TRUE, fig.width=9, eval=FALSE, error= TRUE}
knitr::include_graphics(
  here("_posts/YYYY-MM-DD-post-name/image-name.jpg or png"))
# spacer comment

```


## 3. Save and Knit
Check the resulting HTML in the Viewer window.
Repeat steps 2 (Editing) and 3 (Knitting) until you are happy with the post.

Note that you can keep draft = TRUE in the YAML until you are ready to share with the world


## 4. Sharing
Make sure draft= FALSE in the YAML of your post.

Then Knit it to HTML for a final time.

Then click on the Build Tab, and click on Build Website.

Double check that the HTML looks good before Commit and Push.


## 5. Commit and Push
Click on the Git tab, stage the changed files, Commit them,
and Push to Github.

Wait 1-3 min, then refresh the website at Medical R
https://higgi13425.github.io/medical_r/

to see the results.


## 6. A Warning Note

Building the website will *not* re-knit posts. You have to **choose** to knit any post that you edit.